\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move

### 2025-06-10 writing/appendix-b-gams.Rmd
corrected "log_{10}" in fig cap; need to knit
moved DAGs to main text and updated figures

---
title: "\\Large Predicting how climate change will affect how and where terrestrial mammals will move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure B\!\!} # for "Fig. Bx.
    - \renewcommand{\tablename}{Table B\!\!} # for "Table. Bx.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

@@

# Directed Acyclical Graphs

```{r hgam-dag,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of temperature (Temp) on probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects. Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season. Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality.", fig.height=4, fig.width=4, out.width='40%'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('ggdag')   # for directed acyclical graphs

dagify(Y ~ TOD + DOY + Temp,
       Temp ~ TOD + DOY,
       TOD ~ DOY,
       outcome = 'Y',
       coords = list(x = c(Y = 0, TOD = -1, DOY = 1, Temp = 0),
                     y = c(Y = 1.2, TOD = 2, DOY = 2, Temp = 0))) %>%
  tidy_dagitty() %>%
  mutate(col = case_when(name == 'Y' ~ '1',
                         name == 'TOD' ~ '0',
                         name == 'DOY' ~ '0',
                         name == 'Temp' ~ '2')) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
  geom_dag_point(color = 'black', size = 20, shape = 1) +
  geom_dag_edges() +
  geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
  scale_color_manual(values = c('grey', 'darkorange', 'red')) +
  theme_dag() +
  theme(legend.position = 'none')
```

```{r hrsf-dag, message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of temperature (Temp) on Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W). The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=4, out.width='40%'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('ggdag')   # for directed acyclical graphs

dagify(RSS ~ `F` + E + W,
       `F` ~ Temp,
       E ~ Temp,
       W ~ Temp,
       outcome = 'RSS',
       coords = list(x = c(RSS = 0, `F` = -1, E = 0, W = 1, Temp = 0),
                     y = c(RSS = 2, `F` = 1, E = 1, W = 1, Temp = 0))) %>%
  tidy_dagitty() %>%
  mutate(col = case_when(name == 'RSS' ~ '1',
                         name == 'F' ~ '0',
                         name == 'E' ~ '0',
                         name == 'W' ~ '0',
                         name == 'Temp' ~ '2')) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
  geom_dag_point(color = 'black', size = 20, shape = 1) +
  geom_dag_edges() +
  geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
  scale_color_manual(values = c('grey', 'darkorange', 'red')) +
  theme_dag() +
  theme(legend.position = 'none')
```

@@

```{r dt-smooths, fig.cap="\\textbf{A.} Histograms of sampling intervals between GPS locations, with counts on a $\\log_{10}$ axis for ease of readability. \\textbf{B.} Species-level smooth effects of sampling interval on the estimated probability of the animal moving. \\textbf{C.} Species-level estimated smooth effects of sampling interval on an animal's speed when moving. All x axes are on the natural logarithm scale."}
knitr::include_graphics('../figures/dt-smooths.png')
```

@@

```{r avail-used, out.width='95%', fig.cap="Histograms of available (red) and used (blue) resources. All GPS fixes counts as one observation. The y axis is on the square-root scale to help visualize values with low counts."}
knitr::include_graphics('../figures/available-vs-used.png')
```

>>>>

---
title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure B\!\!} # for "Fig. Bx.
    - \renewcommand{\tablename}{Table B\!\!} # for "Table. Bx.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

@@

@@

```{r dt-smooths, fig.cap="\\textbf{A.} Histograms of sampling intervals between GPS locations. \\textbf{B.} Species-level smooth effects of sampling interval on the estimated probability of the animal moving. \\textbf{C.} Species-level estimated smooth effects of sampling interval on an animal's speed when moving. All x axes are on the natural logarithm scale."}
knitr::include_graphics('../figures/dt-smooths.png')
```

@@

```{r avail-used, out.width='95%', fig.cap="Histograms of available (red) and used (blue) resources. Each GPS fix counts as one observation. The y axis is on the square-root scale to help visualize values with low counts."}
knitr::include_graphics('../figures/available-vs-used.png')
```
